### 2015-01-14 paper.tex
Initial commit for RelEng-2015 paper.

%% This style is provided for the ICSE 2015 main conference,
%% ICSE 2015 co-located events, and ICSE 2015 workshops.

%% bare_conf_ICSE15.tex
%% V1.4
%% 2014/05/22

%% This is a skeleton file demonstrating the use of IEEEtran.cls
%% (requires IEEEtran.cls version 1.7 or later) with an IEEE conference paper.
%%
%% Support sites:
%% http://www.michaelshell.org/tex/ieeetran/
%% http://www.ctan.org/tex-archive/macros/latex/contrib/IEEEtran/
%% and
%% http://www.ieee.org/

%%*************************************************************************
%% Legal Notice:
%% This code is offered as-is without any warranty either expressed or
%% implied; without even the implied warranty of MERCHANTABILITY or
%% FITNESS FOR A PARTICULAR PURPOSE!
%% User assumes all risk.
%% In no event shall IEEE or any contributor to this code be liable for
%% any damages or losses, including, but not limited to, incidental,
%% consequential, or any other damages, resulting from the use or misuse
%% of any information contained here.
%%
%% All comments are the opinions of their respective authors and are not
%% necessarily endorsed by the IEEE.
%%
%% This work is distributed under the LaTeX Project Public License (LPPL)
%% ( http://www.latex-project.org/ ) version 1.3, and may be freely used,
%% distributed and modified. A copy of the LPPL, version 1.3, is included
%% in the base LaTeX documentation of all distributions of LaTeX released
%% 2003/12/01 or later.
%% Retain all contribution notices and credits.
%% ** Modified files should be clearly indicated as such, including  **
%% ** renaming them and changing author support contact information. **
%%
%% File list of work: IEEEtran.cls, IEEEtran_HOWTO.pdf, bare_adv.tex,
%%                    bare_conf.tex, bare_jrnl.tex, bare_jrnl_compsoc.tex
%%*************************************************************************

% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
% *** with production work. IEEE's font choices can trigger bugs that do  ***
% *** not appear when using other class files.                            ***
% The testflow support page is at:
% http://www.michaelshell.org/tex/testflow/

% Note that the a4paper option is mainly intended so that authors in
% countries using A4 can easily print to A4 and see how their papers will
% look in print - the typesetting of the document will not typically be
% affected with changes in paper size (but the bottom and side margins will).
% Use the testflow package mentioned above to verify correct handling of
% both paper sizes by the user's LaTeX system.
%
% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
% should be used if it is desired that the figures are to be displayed in
% draft mode.
%
\documentclass[conference]{IEEEtran}
%
% If IEEEtran.cls has not been installed into the LaTeX system files,
% manually specify the path to it like:
% \documentclass[conference]{../sty/IEEEtran}

% Some very useful LaTeX packages include:
% (uncomment the ones you want to load)

% *** MISC UTILITY PACKAGES ***
%
%\usepackage{ifpdf}
% Heiko Oberdiek's ifpdf.sty is very useful if you need conditional
% compilation based on whether the output is pdf or dvi.
% usage:
% \ifpdf
%   % pdf code
% \else
%   % dvi code
% \fi
% The latest version of ifpdf.sty can be obtained from:
% http://www.ctan.org/tex-archive/macros/latex/contrib/oberdiek/
% Also, note that IEEEtran.cls V1.7 and later provides a builtin
% \ifCLASSINFOpdf conditional that works the same way.
% When switching from latex to pdflatex and vice-versa, the compiler may
% have to be run twice to clear warning/error messages.

% *** CITATION PACKAGES ***
%
%\usepackage{cite}
% cite.sty was written by Donald Arseneau
% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
% \cite{} output to follow that of IEEE. Loading the cite package will
% result in citation numbers being automatically sorted and properly
% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
% \cite will automatically add leading space, if needed. Use cite.sty's
% noadjust option (cite.sty V3.8 and later) if you want to turn this off.
% cite.sty is already installed on most LaTeX systems. Be sure and use
% version 4.0 (2003-05-27) and later if using hyperref.sty. cite.sty does
% not currently provide for hyperlinked citations.
% The latest version can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/cite/
% The documentation is contained in the cite.sty file itself.

% *** GRAPHICS RELATED PACKAGES ***
%
\ifCLASSINFOpdf
  % \usepackage[pdftex]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../pdf/}{../jpeg/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
\else
  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
  % will default to the driver specified in the system graphics.cfg if no
  % driver is specified.
  % \usepackage[dvips]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../eps/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.eps}
\fi
% graphicx was written by David Carlisle and Sebastian Rahtz. It is
% required if you want graphics, photos, etc. graphicx.sty is already
% installed on most LaTeX systems. The latest version and documentation can
% be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/required/graphics/
% Another good source of documentation is "Using Imported Graphics in
% LaTeX2e" by Keith Reckdahl which can be found as epslatex.ps or
% epslatex.pdf at: http://www.ctan.org/tex-archive/info/
%
% latex, and pdflatex in dvi mode, support graphics in encapsulated
% postscript (.eps) format. pdflatex in pdf mode supports graphics
% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
% not a bitmapped formats (.jpeg, .png). IEEE frowns on bitmapped formats
% which can result in "jaggedy"/blurry rendering of lines and letters as
% well as large increases in file sizes.
%
% You can find documentation about the pdfTeX application at:
% http://www.tug.org/applications/pdftex

% *** MATH PACKAGES ***
%
%\usepackage[cmex10]{amsmath}
% A popular package from the American Mathematical Society that provides
% many useful and powerful commands for dealing with mathematics. If using
% it, be sure to load this package with the cmex10 option to ensure that
% only type 1 fonts will utilized at all point sizes. Without this option,
% it is possible that some math symbols, particularly those within
% footnotes, will be rendered in bitmap form which will result in a
% document that can not be IEEE Xplore compliant!
%
% Also, note that the amsmath package sets \interdisplaylinepenalty to 10000
% thus preventing page breaks from occurring within multiline equations. Use:
%\interdisplaylinepenalty=2500
% after loading amsmath to restore such page breaks as IEEEtran.cls normally
% does. amsmath.sty is already installed on most LaTeX systems. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/required/amslatex/math/

% *** SPECIALIZED LIST PACKAGES ***
%
%\usepackage{algorithmic}
% algorithmic.sty was written by Peter Williams and Rogerio Brito.
% This package provides an algorithmic environment fo describing algorithms.
% You can use the algorithmic environment in-text or within a figure
% environment to provide for a floating algorithm. Do NOT use the algorithm
% floating environment provided by algorithm.sty (by the same authors) or
% algorithm2e.sty (by Christophe Fiorio) as IEEE does not use dedicated
% algorithm float types and packages that provide these will not provide
% correct IEEE style captions. The latest version and documentation of
% algorithmic.sty can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithms/
% There is also a support site at:
% http://algorithms.berlios.de/index.html
% Also of interest may be the (relatively newer and more customizable)
% algorithmicx.sty package by Szasz Janos:
% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithmicx/

% *** ALIGNMENT PACKAGES ***
%
%\usepackage{array}
% Frank Mittelbach's and David Carlisle's array.sty patches and improves
% the standard LaTeX2e array and tabular environments to provide better
% appearance and additional user controls. As the default LaTeX2e table
% generation code is lacking to the point of almost being broken with
% respect to the quality of the end results, all users are strongly
% advised to use an enhanced (at the very least that provided by array.sty)
% set of table tools. array.sty is already installed on most systems. The
% latest version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/required/tools/

%\usepackage{mdwmath}
%\usepackage{mdwtab}
% Also highly recommended is Mark Wooding's extremely powerful MDW tools,
% especially mdwmath.sty and mdwtab.sty which are used to format equations
% and tables, respectively. The MDWtools set is already installed on most
% LaTeX systems. The lastest version and documentation is available at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/mdwtools/

% IEEEtran contains the IEEEeqnarray family of commands that can be used to
% generate multiline equations as well as matrices, tables, etc., of high
% quality.

%\usepackage{eqparbox}
% Also of notable interest is Scott Pakin's eqparbox package for creating
% (automatically sized) equal width boxes - aka "natural width parboxes".
% Available at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/eqparbox/

% *** SUBFIGURE PACKAGES ***
%\usepackage[tight,footnotesize]{subfigure}
% subfigure.sty was written by Steven Douglas Cochran. This package makes it
% easy to put subfigures in your figures. e.g., "Figure 1a and 1b". For IEEE
% work, it is a good idea to load it with the tight package option to reduce
% the amount of white space around the subfigures. subfigure.sty is already
% installed on most LaTeX systems. The latest version and documentation can
% be obtained at:
% http://www.ctan.org/tex-archive/obsolete/macros/latex/contrib/subfigure/
% subfigure.sty has been superceeded by subfig.sty.

%\usepackage[caption=false]{caption}
%\usepackage[font=footnotesize]{subfig}
% subfig.sty, also written by Steven Douglas Cochran, is the modern
% replacement for subfigure.sty. However, subfig.sty requires and
% automatically loads Axel Sommerfeldt's caption.sty which will override
% IEEEtran.cls handling of captions and this will result in nonIEEE style
% figure/table captions. To prevent this problem, be sure and preload
% caption.sty with its "caption=false" package option. This is will preserve
% IEEEtran.cls handing of captions. Version 1.3 (2005/06/28) and later
% (recommended due to many improvements over 1.2) of subfig.sty supports
% the caption=false option directly:
%\usepackage[caption=false,font=footnotesize]{subfig}
%
% The latest version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/subfig/
% The latest version and documentation of caption.sty can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/caption/

% *** FLOAT PACKAGES ***
%
%\usepackage{fixltx2e}
% fixltx2e, the successor to the earlier fix2col.sty, was written by
% Frank Mittelbach and David Carlisle. This package corrects a few problems
% in the LaTeX2e kernel, the most notable of which is that in current
% LaTeX2e releases, the ordering of single and double column floats is not
% guaranteed to be preserved. Thus, an unpatched LaTeX2e can allow a
% single column figure to be placed prior to an earlier double column
% figure. The latest version and documentation can be found at:
% http://www.ctan.org/tex-archive/macros/latex/base/

%\usepackage{stfloats}
% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
% the ability to do double column floats at the bottom of the page as well
% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
% LaTeX2e). It also provides a command:
%\fnbelowfloat
% to enable the placement of footnotes below bottom floats (the standard
% LaTeX2e kernel puts them above bottom floats). This is an invasive package
% which rewrites many portions of the LaTeX2e float routines. It may not work
% with other packages that modify the LaTeX2e float routines. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/sttools/
% Documentation is contained in the stfloats.sty comments as well as in the
% presfull.pdf file. Do not use the stfloats baselinefloat ability as IEEE
% does not allow \baselineskip to stretch. Authors submitting work to the
% IEEE should note that IEEE rarely uses double column equations and
% that authors should try to avoid such use. Do not be tempted to use the
% cuted.sty or midfloat.sty packages (also by Sigitas Tolusis) as IEEE does
% not format its papers in such ways.

% *** PDF, URL AND HYPERLINK PACKAGES ***
%
%\usepackage{url}
% url.sty was written by Donald Arseneau. It provides better support for
% handling and breaking URLs. url.sty is already installed on most LaTeX
% systems. The latest version can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/misc/
% Read the url.sty source comments for usage information. Basically,
% \url{my_url_here}.

% *** Do not adjust lengths that control margins, column widths, etc. ***
% *** Do not use packages that alter fonts (such as pslatex).         ***
% There should be no need to do such things with IEEEtran.cls V1.6 and later.
% (Unless specifically asked to do so by the journal or conference you plan
% to submit to, of course. )

\usepackage{color}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{url}

% correct bad hyphenation here
\hyphenation{op-tical net-works semi-conduc-tor}

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{Continuous Deployment and Schema Evolution \\in SQL Databases}

% author names and affiliations
% use a multiple column layout for up to three different
% affiliations
%\author{\IEEEauthorblockN{Michael Shell}
%\IEEEauthorblockA{School of Electrical and\\Computer Engineering\\
%Georgia Institute of Technology\\
%Atlanta, Georgia 30332--0250\\
%Email: http://www.michaelshell.org/contact.html}
%\and
%\IEEEauthorblockN{Homer Simpson}
%\IEEEauthorblockA{Twentieth Century Fox\\
%Springfield, USA\\
%Email: homer@thesimpsons.com}
%\and
%\IEEEauthorblockN{James Kirk\\ and Montgomery Scott}
%\IEEEauthorblockA{Starfleet Academy\\
%San Francisco, California 96678-2391\\
%Telephone: (800) 555--1212\\
%Fax: (888) 555--1212}}

% conference papers do not typically use \thanks and this command
% is locked out in conference mode. If really needed, such as for
% the acknowledgment of grants, issue a \IEEEoverridecommandlockouts
% after \documentclass

% for over three affiliations, or if they all won't fit within the width
% of the page, use this alternative format:
%
\author{
  \IEEEauthorblockN{
    Michael de Jong\IEEEauthorrefmark{1},
    Arie van Deursen\IEEEauthorrefmark{2}
  }
  \IEEEauthorblockA{
    Delft University of Technology,\\
    Delft, Netherlands\\ \\
    Email: M.deJong-2@student.tudelft.nl\IEEEauthorrefmark{1}\\
    Email: Arie.vanDeursen@tudelft.nl\IEEEauthorrefmark{2}\\ \\
    \today{}
  }
}

% use for special paper notices
%\IEEEspecialpapernotice{(Invited Paper)}

% make the title area
\maketitle

\begin{abstract}
%\boldmath
The abstract goes here.

\end{abstract}
% IEEEtran.cls defaults to using nonbold math in the Abstract.
% This preserves the distinction between vectors and scalars. However,
% if the conference you are submitting to favors bold math in the abstract,
% then you can use LaTeX's standard command \boldmath at the very start
% of the abstract to achieve this. Many IEEE journals/conferences frown on
% math in the abstract anyway.

% no keywords

% For peer review papers, you can put extra information on the cover
% page as needed:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi
%
% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
\IEEEpeerreviewmaketitle

\section{Introduction} %(0.7)

% * Problem from RelEng viewpoint.
% * Blocking nature of DDL statements.
% * Continuous Deployment and partial deployments.

SQL databases are an integral part of many (web) applications. They provide strong guarantees about the storage and retrieval of data which makes them useful for various software applications. However the environment of software engineers and release engineers is drastically changing: shorter development iterations, more frequently deployed to production, or even automatically deployed to production several times a day. For release engineers this brings additional challenges. In particular when evolving the schema (tables, columns, etc) of the SQL database.

This can essentially be distilled into two separate problems:
\begin{itemize}
  \item{Given the deployment of a new version of a web application, depending on the deployment strategy it might be required that both the old and the new version of the application run in parallel (\textit{Rolling Upgrades}, or \textit{Canary Releases}). This means that the schema on the SQL database needs to be compatible with the schema expected by both the old and the new application.}
  \item{Operations which modify the schema in SQL databases (\textit{DDL statements}), depending on the SQL database implementation may be either non-blocking, or blocking in nature. In case of non-blocking behaviour, queries issued by the application for instance may proceed, whereas with blocking behaviour these queries may be put on hold for the duration of the schema modification. This can range anywhere from milliseconds to hours depending on the operation, the size of the affected table, and database implementation.}
\end{itemize}

These two problems make releasing often, problematic for applications with no service windows. To address this problem we aim to implement our own proof-of-concept tooling, which would allow its users to not only version their database schema, but also evolve their database schema without affecting older clients because of changes in the database schema or using blocking schema operations.

\section{State of the art} %(0.7)

% * Manual approach
% * Automated tools in academia and industry
% * Limitations of approaches

Before we decided to develop our own tooling, we examined existing approaches, tools and research which aims to solve these problems. A good starting point is the seminal book ``Continuous Delivery''~\cite{Humble:2010:CDR:1869904}. This book describes three high-level approaches to deal with migrating data without requiring downtime:

\begin{itemize}
  \item{\textbf{Approach 1}: Copy the database and start recording transactions. When the copying has completed, apply the schema changes. Finally replay all the recorded database transactions to reach an equivalent database.}
  \item{\textbf{Approach 2}: When using Blue-Green deployments, one possible approach is to copy the database to the alternative production environment and apply the schema operations there. Meanwhile these two databases need to be kept in sync, or the original must be in a read-only state.}
  \item{\textbf{Approach 3}: Another approach is performing schema operations separately from your application deployments using operations which are non-blocking. As a consequence the application must be able to deal with either the old or the new database schema being in use, and formulate queries accordingly.}
\end{itemize}

\subsection{Current tooling} 
Since evolving the schema of a database and coordinating the deployment of software is a complex task, we looked for tooling which could help automate these tasks for developers and release engineers.

\subsubsection{Openark Kit, Percona Toolkit}
The Openark Kit~\cite{OpenarkKit} and Percona Toolkit~\cite{PerconaToolkit} are tools designed to help DBAs perform administrative tasks on MySQL databases. \textit{oak-online-alter-table} and \textit{pt-online-schema-change} respectively are commands which follow a variation of approach 1. They create an empty copy of the a specified table and then apply schema operations on the new table. These commands then proceed to install triggers on both tables, which insert, update, or delete the corresponding record in the other table when a record in their own table is inserted, updated, or deleted. A background task then iterates over all rows in the original table (in batches) and lazily copies rows from the original table to the new table. When all rows have been copied the original table and new table switch names, effectively ensuring that clients now use the new table when querying the old table name.

\subsubsection{TableMigrator}
TableMigrator~\cite{TableMigrator} is a tool which works very similar to Openark Kit and Percona Toolkit. But instead of using triggers to synchronize the two tables with each other. TableMigrator requires tables to have an additional column which stores the date and time that record was last modified. Using this information corresponding rows in both tables are synchronized with each other by passing over all rows in multiple passes. Once a pass can be done relatively quickly, TableMigrator acquires an exclusive write lock --- blocking the entire table --- and proceeds to do a final pass. Once this has been done, it atomically swaps both table names before releasing the write lock.

\subsubsection{Online Schema Change \& Large Hadron Migrator}
Facebook's Online Schema Change (\textit{OSC})~\cite{FacebookOSC} and Soundcloud's Large Hadron Migrator (\textit{LHM})~\cite{SoundcloudLHM} are both tools which closely follow approach 1. These tools create a full copy of a table (including its contents) and record modifications to rows in the original table using triggers into a special \textit{deltas} table. Once they have copied the original table, they apply the specified schema operations to the new table before proceeding to replay the recorded modifications from the \textit{deltas} table onto the new table. When the \textit{deltas} table is almost empty, these tools acquire an exclusive write lock, apply the last recorded modifications and then atomically switch the names of both tables.

\subsection{Research}

\subsubsection{PRISM?}
\textbf{Should I cover PRISM and SMOs here?}

\subsubsection{Imago}
Imago is a tool which follows approach 2. With Imago there are two production environments --- called "universes" by the authors. Imago integrates at both ingress point (the load balancer) to be able to switch users from one universe to the other, and at the egress point where the (web) application sends queries to the SQL database. When a new version of the application (requiring a schema change) needs to be deployed, Imago bootstraps the inactive production environment with a copy of the database, applies the required schema operation and then copies over the data from the old universe to the new universe. When this process is almost complete, it enforces a short period of read-only access to the SQL database while it is switching users from the old environment to the new using the load balancer.

\subsection{Limitations}

\begin{itemize}
  \item{None of the tools used in practice support foreign keys and primarily focus on MySQL databases. Imago makes no claims with regards to support for foreign keys, but does impose restrictions on how to deploy applications, and makes assumptions on the read/write ratio of database queries.}
  \item{Tools which do not handle the deployment of the application themselves, also force the developers of the application to ensure that their application is able to deal with the schema being in either the old or the new state. All tools atomically rename the modified table, but the application itself is left to deal with modifying queries to accommodate added or removed columns, constraints, or tables at the time of the atomic table rename operation.}
\end{itemize}

\section{Our approach} %(1.0)

\subsection{Vision}

We feel that the current situation is an impediment to adoption and growth of \textit{Continuous Deployment}. Although \textit{Continuous Deployment} has made great improvements on workflows, team dynamics, and tooling for deploying and monitoring applications, the area of schema evolution in databases has received little attention. Therefore we aim to address these challenges by developing our own tool which adheres to the following principles:

\begin{itemize}
  \item{Do not restrict release engineers, or tell them how to deploy their applications. Release engineers have their own tools to perform deployments.}
  \item{If an application requires a SQL database, it should be exposed as one or more database schemas simultaneously which are synchronized with each other in a transactionally safe manner. Each client should connect to the database using one particular schema. This allows for \textit{Canary/Feature Deployments}, as well as \textit{Blue-Green deployments} and \textit{Rolling upgrades}.}
  \item{Versioning and rolling back are important for release engineers. Any tool which modifies the database schema should allow for versioning and be able to rollback to previous versions.}
  \item{Foreign keys are an integral part of SQL databases and should be supported.}
\end{itemize}

\subsection{Approach}

\section{Implementation status} %(0.5)

\begin{itemize}
  \item{Current state of implementation}
  \item{Road ahead?}
\end{itemize}

\section{Early evaluation} %(0.5)

\begin{itemize}
  \item{Experiments at Magnet.me}
  \item{Figures}
\end{itemize}

% An example of a floating figure using the graphicx package.
% Note that \label must occur AFTER (or within) \caption.
% For figures, \caption should occur after the \includegraphics.
% Note that IEEEtran v1.7 and later has special internal code that
% is designed to preserve the operation of \label within \caption
% even when the captionsoff option is in effect. However, because
% of issues like this, it may be the safest practice to put all your
% \label just after \caption rather than within \caption{}.
%
% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
% option should be used if it is desired that the figures are to be
% displayed while in draft mode.
%
%\begin{figure}[!t]
%\centering
%\includegraphics[width=2.5in]{myfigure}
% where an .eps filename suffix will be assumed under latex,
% and a .pdf suffix will be assumed for pdflatex; or what has been declared
% via \DeclareGraphicsExtensions.
%\caption{Simulation Results}
%\label{fig_sim}
%\end{figure}

% Note that IEEE typically puts floats only at the top, even when this
% results in a large percentage of a column being occupied by floats.

% An example of a double column floating figure using two subfigures.
% (The subfig.sty package must be loaded for this to work.)
% The subfigure \label commands are set within each subfloat command, the
% \label for the overall figure must come after \caption.
% \hfil must be used as a separator to get equal spacing.
% The subfigure.sty package works much the same way, except \subfigure is
% used instead of \subfloat.
%
%\begin{figure*}[!t]
%\centerline{\subfloat[Case I]\includegraphics[width=2.5in]{subfigcase1}%
%\label{fig_first_case}}
%\hfil
%\subfloat[Case II]{\includegraphics[width=2.5in]{subfigcase2}%
%\label{fig_second_case}}}
%\caption{Simulation results}
%\label{fig_sim}
%\end{figure*}
%
% Note that often IEEE papers with subfigures do not employ subfigure
% captions (using the optional argument to \subfloat), but instead will
% reference/describe all of them (a), (b), etc., within the main caption.

% An example of a floating table. Note that, for IEEE style tables, the
% \caption command should come BEFORE the table. Table text will default to
% \footnotesize as IEEE normally uses this smaller font for tables.
% The \label must come after \caption as always.
%
%\begin{table}[!t]
%% increase table row spacing, adjust to taste
%\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
%\caption{An Example of a Table}
%\label{table_example}
%\centering
%% Some packages, such as MDW tools, offer better commands for making tables
%% than the plain LaTeX2e tabular which is used here.
%\begin{tabular}{|c||c|}
%\hline
%One & Two\\
%\hline
%Three & Four\\
%\hline
%\end{tabular}
%\end{table}

% Note that IEEE does not put floats in the very first column - or typically
% anywhere on the first page for that matter. Also, in-text middle ("here")
% positioning is not used. Most IEEE journals/conferences use top floats
% exclusively. Note that, LaTeX2e, unlike IEEE journals/conferences, places
% footnotes above bottom floats. This can be corrected via the \fnbelowfloat
% command of the stfloats package.

\section{Conclusion} 
The conclusion goes here.

% conference papers do not normally have an appendix

% use section* for acknowledgement
%\section*{Acknowledgment}

%The authors would like to thank...

% trigger a \newpage just before the given reference
% number - used to balance the columns on the last page
% adjust value as needed - may need to be readjusted if
% the document is modified later
%\IEEEtriggeratref{8}
% The "triggered" command can be changed if desired:
%\IEEEtriggercmd{\enlargethispage{-5in}}

% references section

% can use a bibliography generated by BibTeX as a .bbl file
% BibTeX documentation can be easily obtained at:
% http://www.ctan.org/tex-archive/biblio/bibtex/contrib/doc/
% The IEEEtran BibTeX style support page is at:
% http://www.michaelshell.org/tex/ieeetran/bibtex/
%\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
%\bibliography{IEEEabrv,../bib/paper}
%
% <OR> manually copy in the resultant .bbl file
% set second argument of \begin to the number of references
% (used to reserve space for the reference number labels box)

\bibliographystyle{plain}
\bibliography{paper}

% \begin{thebibliography}{1}
% 
% \bibitem{IEEEhowto:kopka}
% H.~Kopka and P.~W. Daly, \emph{A Guide to \LaTeX}, 3rd~ed.\hskip 1em plus
%   0.5em minus 0.4em\relax Harlow, England: Addison-Wesley, 1999.
% 
% \end{thebibliography}

% that's all folks
\end{document}

### 2015-01-21 paper.tex
Improved draft, added 'approach' and 'implementation status' sections.

@@
 %
-%\usepackage{cite}
+\usepackage{cite}
 % cite.sty was written by Donald Arseneau
@@
 %
-%\usepackage{url}
+\usepackage{url}
 % url.sty was written by Donald Arseneau. It provides better support for
@@
 % to submit to, of course. )
-
-\usepackage{color}
-\usepackage[usenames,dvipsnames]{xcolor}
-\usepackage{url}
 
@@
 For release engineers this brings additional challenges.
-In particular when evolving the schema (tables, columns, etc) of the SQL database.
+In particular when evolving the schema (tables, columns, foreign keys, etc) of the SQL database either forwards or backwards without loss of data.
 
@@
   \item{Operations which modify the schema in SQL databases (\textit{DDL statements}), depending on the SQL database implementation may be either non-blocking, or blocking in nature.
-In case of non-blocking behaviour, queries issued by the application for instance may proceed, whereas with blocking behaviour these queries may be put on hold for the duration of the schema modification.
-This can range anywhere from milliseconds to hours depending on the operation, the size of the affected table, and database implementation.}
+In case of non-blocking DDL statements queries issued by the application for instance may proceed, whereas with blocking DDL statements these queries may be put on hold for the duration of the schema modification.
+This disruption can range anywhere from milliseconds to hours depending on the DDL statement, the size of the affected table, and database implementation.
+One such example would be adding a new non-nullable column to a PostgreSQL database.
+This operation is blocking in nature --- preventing other queries from running while it is being executed --- and when applied to a table of 100 million rows (in our tests) blocked other queries from running on that table for 5 minutes.}
 \end{itemize}
 
-These two problems make releasing often, problematic for applications with no service windows.
-To address this problem we aim to implement our own proof-of-concept tooling, which would allow its users to not only version their database schema, but also evolve their database schema without affecting older clients because of changes in the database schema or using blocking schema operations.
+When service windows are unavailable, these two problems make releasing applications often, problematic for release engineers.
+To address this problem we aim to find or implement tooling which would allow its users to not only version their database schema, but also evolve their database schema without affecting older and still running applications.
 
@@
 
-Before we decided to develop our own tooling, we examined existing approaches, tools and research which aims to solve these problems.
+Before we decided to develop tooling to solve this ourselves, we examined existing approaches, tools, and research which aims to solve these problems.
 A good starting point is the seminal book ``Continuous Delivery''~\cite{Humble:2010:CDR:1869904}.
-This book describes three high-level approaches to deal with migrating data without requiring downtime:
+This book describes three high-level approaches to deal with migrating data without data loss or requiring downtime:
 
 \begin{itemize}
-  \item{\textbf{Approach 1}: Copy the database and start recording transactions.
-When the copying has completed, apply the schema changes.
-Finally replay all the recorded database transactions to reach an equivalent database.}
-  \item{\textbf{Approach 2}: When using Blue-Green deployments, one possible approach is to copy the database to the alternative production environment and apply the schema operations there.
-Meanwhile these two databases need to be kept in sync, or the original must be in a read-only state.}
-  \item{\textbf{Approach 3}: Another approach is performing schema operations separately from your application deployments using operations which are non-blocking.
+  \item{\textbf{Approach 1}: Copy the database or schema and start recording transactions.
+When the copying has completed, apply the schema changes to the copied database or schema.
+Finally replay all the recorded database transactions on the copied database or schema to reach an equivalent state.
+Once this state has been achieved, it must either be maintained by continuing to record and playback transaction, or one must make the switch from the old to the new database or schema.}
+  \item{\textbf{Approach 2}: When using Blue-Green deployments, two instances of the database are available.
+In order to evolve the schema one must apply the schema operations to the inactive database.
+In the mean time these two databases need to be kept synchronized.
+This can be done by using approach 1, or somehow ensure that the table under change remains in a read-only state until the active and inactive databases have switched roles.}
+  \item{\textbf{Approach 3}: The last is an approach where schema operations are performed separately from your application deployments using operations which are non-blocking.
 As a consequence the application must be able to deal with either the old or the new database schema being in use, and formulate queries accordingly.}
@@
 
-\subsubsection{Openark Kit, Percona Toolkit}
-The Openark Kit~\cite{OpenarkKit} and Percona Toolkit~\cite{PerconaToolkit} are tools designed to help DBAs perform administrative tasks on MySQL databases.
+\subsubsection{Percona Toolkit~\cite{PerconaToolkit} \& Openark Kit~\cite{OpenarkKit}}
+are tools designed to help DBAs perform administrative tasks on MySQL databases.
 \textit{oak-online-alter-table} and \textit{pt-online-schema-change} respectively are commands which follow a variation of approach 1.
@@
 
-\subsubsection{TableMigrator}
-TableMigrator~\cite{TableMigrator} is a tool which works very similar to Openark Kit and Percona Toolkit.
+\subsubsection{TableMigrator~\cite{TableMigrator}}
+is a tool which works very similar to Openark Kit and Percona Toolkit.
 But instead of using triggers to synchronize the two tables with each other.
@@
 
-\subsubsection{Online Schema Change \& Large Hadron Migrator}
-Facebook's Online Schema Change (\textit{OSC})~\cite{FacebookOSC} and Soundcloud's Large Hadron Migrator (\textit{LHM})~\cite{SoundcloudLHM} are both tools which closely follow approach 1.
+\subsubsection{SoundCloud's Large Hadron Migrator~\cite{SoundcloudLHM} \& Facebook's Online Schema Change~\cite{FacebookOSC}}
+are both tools which closely follow approach 1.
 These tools create a full copy of a table (including its contents) and record modifications to rows in the original table using triggers into a special \textit{deltas} table.
@@
 
-\subsubsection{PRISM?}
-\textbf{Should I cover PRISM and SMOs here?}
+% Technically we could go without this subsubsection...
+\subsubsection{PRISM}
+PRISM~\cite{Curino:2010:URI:1921071.1921078, Curino:2009:ADS:1656437.1656444} is a tool which is a part of a larger framework.
+Although PRISM does not solve the problem of downtime when performing operations on the database schema, it does aim to shorten it.
+PRISM is designed to reason about the database schema changes, and infer how to rewrite existing queries to the new resulting database schema.
+In addition it can generate the migration and rollback scripts which will transform the current database schema to the new database schema and back.
+The most influential concept of PRISM is the definition of several new operations to evolve a database schema.
+These described Schema Modification Operations (\textit{SMOs}), offer ways to split or merge tables into multiple tables (structurally or content-wise).
 
 \subsubsection{Imago}
-Imago is a tool which follows approach 2.
+Imago~\cite{Dumitras:2009:WUF:1656980.1657005} is a tool which follows approach 2.
 With Imago there are two production environments --- called "universes" by the authors.
@@
 When a new version of the application (requiring a schema change) needs to be deployed, Imago bootstraps the inactive production environment with a copy of the database, applies the required schema operation and then copies over the data from the old universe to the new universe.
-When this process is almost complete, it enforces a short period of read-only access to the SQL database while it is switching users from the old environment to the new using the load balancer.
+When this process is almost complete, it enforces a short period of read-only access to the SQL database while it is switching users from the old environment to the new, using the load balancer.
+Imago makes no claims in regards to its support for foreign keys.
 
@@
 \begin{itemize}
-  \item{None of the tools used in practice support foreign keys and primarily focus on MySQL databases.
-Imago makes no claims with regards to support for foreign keys, but does impose restrictions on how to deploy applications, and makes assumptions on the read/write ratio of database queries.}
-  \item{Tools which do not handle the deployment of the application themselves, also force the developers of the application to ensure that their application is able to deal with the schema being in either the old or the new state.
-All tools atomically rename the modified table, but the application itself is left to deal with modifying queries to accommodate added or removed columns, constraints, or tables at the time of the atomic table rename operation.}
+  \item{None of the tools used in practice support foreign keys and primarily focus on MySQL databases.}
+  \item{Tools which do not handle the deployment of the application themselves, all force the developers of the application to ensure that their application is able to deal with the database schema being in either the old or the new state.
+All tools atomically rename the modified table, but the software developers of the application are left to deal with modifying queries to accommodate added or removed columns, constraints, or tables at the time of the atomic table rename operation.}
 \end{itemize}
@@
 We feel that the current situation is an impediment to adoption and growth of \textit{Continuous Deployment}.
-Although \textit{Continuous Deployment} has made great improvements on workflows, team dynamics, and tooling for deploying and monitoring applications, the area of schema evolution in databases has received little attention.
-Therefore we aim to address these challenges by developing our own tool which adheres to the following principles:
+Although \textit{Continuous Deployment} has made great improvements on workflows, team dynamics, and tooling for deploying and monitoring applications, the area of schema evolution in databases has received little attention and no suitable solution currently exists.
+Therefore we aim to address these challenges by developing a new tool which adheres to the following restrictions:
 
 \begin{itemize}
-  \item{Do not restrict release engineers, or tell them how to deploy their applications. Release engineers have their own tools to perform deployments.}
-  \item{If an application requires a SQL database, it should be exposed as one or more database schemas simultaneously which are synchronized with each other in a transactionally safe manner.
-Each client should connect to the database using one particular schema.
-This allows for \textit{Canary/Feature Deployments}, as well as \textit{Blue-Green deployments} and \textit{Rolling upgrades}.}
-  \item{Versioning and rolling back are important for release engineers. Any tool which modifies the database schema should allow for versioning and be able to rollback to previous versions.}
+  \item{A solution should not put any restrictions on how to deploy the application, or which methods of deployment that can be used: \textit{Blue-Green deployments}, \textit{Rolling upgrades}, \textit{Canary releases}, etc.}
+  \item{The contents of the database must be accessible through one or more database schemas, which are continuously kept synchronized with each other.}
+  \item{Clients connected to the database must be able to ``view'' the contents of the database according to an active database schema of their choice.
+Their ``view'' should only contain tables, columns, foreign keys, etc which are present in their chosen database schema.
+Even when more than one database schema is active at the same time.}
+  \item{Versioning and rolling back are important for release engineers.
+A solution should support versioning the changes being made to the database schema, and be able to rollback to a previous version.}
   \item{Foreign keys are an integral part of SQL databases and should be supported.}
@@
 
-\subsection{Approach}
+\subsection{Solution direction}
+
+We have come up with an approach which combines approaches 1 and 3.
+The previously discussed tools from practice all use approach 1, and prove that this is a viable way for performing simple evolutions of the database schema.
+These tools all create a single ghost table, whereas we aim to create one ghost table for each table that is under change, and ghost tables for all tables which in turn are dependent on these tables under change.
+This approach ensures that we do not violate any previously defined foreign keys.
+
+\textbf{Show image an example here?}
+
+In addition, the tools from practice all allow release engineers to switch the table names manually after the the original and the ghost table have been synchronized.
+This is good since it allows for two or more versions of the application to co-exist on the same database as long as each version correctly queries the correct (ghost)table.
+Although the current tools from practice do not solve this specific problem, we think that this is easily solved using a small wrapper around database drivers for most programming languages.
+Alternatively, for software engineers working on the application we could expose which (ghost)table should be queried for a specific version of the database schema, when writing SQL queries.
 
@@
 
+\subsection{Current status}
+
+We are implementing a new tool --- QuantumDB~\cite{QuantumDB} --- which aims to address the need for non-blocking schema evolution, and remove obstacles for teams who want to adopt \textit{Continuous Deployment} but struggle because of their SQL database(s).
+Although not yet usable in production environments, we have implemented enough features at the moment to take a step back and evaluate QuantumDB in an small experiment.
+
+QuantumDB currently consists of two parts: 
 \begin{itemize}
-  \item{Current state of implementation}
-  \item{Road ahead?}
+  \item{\textbf{core} is the program responsible for performing the actual migration tasks on the database.
+It does this by creating ghost tables of tables under change, using triggers to keep the ghost tables and the original tables synchronized with each other, and a background task which copies the data from the original tables to the ghost tables in small batches to prevent overloading, and degrading the performance of the database.}
+  \item{\textbf{driver} is a Java JDBC driver which can wrap another JDBC driver.
+When used, the application developers can specify a schema version on which their application operates.
+The driver then rewrites SQL queries sent to it by the application, and passes them on to the wrapped database JDBC driver which in turn sends it to the database.
+When rewriting queries, it replaces table names with the names of the associated (ghost)table names if applicable.}
+\end{itemize}
+
+With QuantumDB we can currently take one of the active database schemas, apply an arbitrary number of schema operations to it, and expose the resulting database schema in addition to the original database schemas.
+Once none of the applications are still connected to a particular version of the database schema, it can safely be discarded in a background task.
+Effectively this can be described as A/B testing the database schema.
+
+\subsection{Road ahead}
+
+As stated before, the implementation of QuantumDB is not yet ready for production. To be able to use QuantumDB effectively in production environments, we still need to implement some features.
+
+\begin{itemize}
+  \item{QuantumDB can already deal with foreign keys referring from tables under change to other tables not under change.
+To fully support foreign keys QuantumDB would also need to be able to deal with foreign keys referring to tables under change.}
+  \item{QuantumDB currently only supports PostgreSQL, but should prove easily extensible to other popular SQL databases since most of these have support for triggers and procedures.}
+  \item{The driver currently uses a very rudimentary query rewriting algorithm to replace table names under change with the names of ghost tables.
+It does not yet support the entire PostgreSQL query syntax.}
+  % Technically speaking this next one is optional.%
+  \item{Adding support for SMOs. These operations allow for more complex schema operations on tables, and should enable software engineers to make more elaborate changes to the database schema.}
+  \item{Evaluate QuantumDB more extensively in use cases at different organizations and in production environments.}
 \end{itemize}
@@
 \begin{itemize}
-  \item{Experiments at Magnet.me}
-  \item{Figures}
+  \item{Experiment setup}
+  \item{Results}
 \end{itemize}

### 2015-01-23 paper.tex
Added the 'initial measurements' section, conclusion and abstract.

@@
 \ifCLASSINFOpdf
-  % \usepackage[pdftex]{graphicx}
+  \usepackage[pdftex]{graphicx}
+  \usepackage{subfig}
   % declare the path(s) where your graphic files are
@@
   % every instance of \includegraphics
-  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
+  \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
 \else
@@
 %
+
 \author{
@@
 %\boldmath
-The abstract goes here.
+
+Continuous Deployment is a hot topic.
+It holds much promise for developing and deploying software applications, yet both software engineers and release engineers face various challenges when adopting this practice.
+In our view, the biggest technical and operational challenge is evolving the database schema of SQL databases without disrupting the clients connected to the database.
+It also can have an impact on how the application is developed, since big schema changes cannot be deployed without downtime, requiring intermediate versions of the application.
+In this paper we explore tools, and approaches to address this problem.
 
@@
 One such example would be adding a new non-nullable column to a PostgreSQL database.
-This operation is blocking in nature --- preventing other queries from running while it is being executed --- and when applied to a table of 100 million rows (in our tests) blocked other queries from running on that table for 5 minutes.}
+This operation is blocking in nature --- preventing other queries from running while it is being executed --- and when applied to a table of 100 million rows (in our tests) blocked other queries from running on that table for 2.5 minutes.}
 \end{itemize}
@@
 \subsection{Research}
-
-% Technically we could go without this subsubsection...
-\subsubsection{PRISM}
-PRISM~\cite{Curino:2010:URI:1921071.1921078, Curino:2009:ADS:1656437.1656444} is a tool which is a part of a larger framework.
-Although PRISM does not solve the problem of downtime when performing operations on the database schema, it does aim to shorten it.
-PRISM is designed to reason about the database schema changes, and infer how to rewrite existing queries to the new resulting database schema.
-In addition it can generate the migration and rollback scripts which will transform the current database schema to the new database schema and back.
-The most influential concept of PRISM is the definition of several new operations to evolve a database schema.
-These described Schema Modification Operations (\textit{SMOs}), offer ways to split or merge tables into multiple tables (structurally or content-wise).
 
@@
 
-\subsection{Road ahead}
+\section{Initial measurements} %(0.5)
 
+We feel it is important to validate our approach at an early stage.
+Although QuantumDB is far from ready, we setup a small experiment to verify that the concept and our implementation of it works, and that it can match the capabilities of the tools offered in practice.
+
+\subsection{Setup}
+We have installed PostgreSQL 9.3 and a test application --- available on GitHub~\cite{QuantumDB-RelEng-Demo} --- on a server with a 4-core Intel Core i7-3770, a 1TB hard disk, and 16GB memory.
+The test application inserts 100 million records into a single table called ``users'' and starts a simple demo application with two threads which continuously query and manipulate the data in that table using SELECT, INSERT, UPDATE, and DELETE statements.
+It then uses one of two methods to add a non nullable column to the ``users'' table.
+During this process we keep track of the performance of the queries issued by the demo application and how long each method takes to complete.
+
+The ``naive'' approach simply executes an ``\textit{ALTER TABLE users ADD COLUMN}'' query to add the new column to the database.
+Once this query has completed, we can start a new demo application which can operate on the new structure of the ``users'' table.
+Since the new structure of the ``users'' table is compatible with both the old and the new version of the demo application they can run in parallel to each other, allowing for any kind of deployment method to be used.
+
+Our approach creates a ghost table like much the tools from practice do.
+This ghost table is created with the new structure of the ``users'' table, and then filled iteratively in batches of 2,000 records at intervals of 50 milliseconds in an attempt to avoid negatively affecting the performance of the queries issued by the demo application.
+Unlike with the previous approach, with this approach the demo applications use the QuantumDB JDBC driver which expose each demo application to a different database schema (containing either the ``users'' table or the ghost table), while QuantumDB core keeps both tables sychronized to each other using database triggers.
+
+\subsection{Results}
+
+The total migration time for the ``naive'' approach was 2 minutes and 26 seconds, during which the demo application was unable to read from or write to the database.
+The total migration time for the QuantumDB approach was 50 minutes and 41 seconds.
+Although significantly slower than the ``naive'' approach, the demo applications remained operational throughout this period (see Figure~\ref{fig:migration}).
+
+\begin{figure}%
+    \centering
+    \subfloat[Before migration]{{\includegraphics[width=1.55in]{normal-operations.png} }}%
+    \qquad
+    \subfloat[During migration]{{\includegraphics[width=1.55in]{migration.png} }}%
+    \caption{Query performance of demo application before and during migration}%
+    \label{fig:migration}%
+\end{figure}
+
+It is worth noting, that we have not invested any time in improving the performance of QuantumDB or reducing the performance impact it has on the database and database clients.
+During this experiment we noticed that the hard disk was periodically unable to write data to disk quickly enough, resulting in queries taking longer to complete than regularly.
+We are confident that we can greatly reduce this problem by tweaking the PostgreSQL config, or switching to a faster hard disk.
+
+\section{Road ahead}
 As stated before, the implementation of QuantumDB is not yet ready for production. To be able to use QuantumDB effectively in production environments, we still need to implement some features.
@@
 To fully support foreign keys QuantumDB would also need to be able to deal with foreign keys referring to tables under change.}
+  \item{Use metrics to throttle the copy phase.
+When the database is (too) busy we can slow down the iterative copy process, and vice-versa speed it up when the database is not under significant load.}
   \item{QuantumDB currently only supports PostgreSQL, but should prove easily extensible to other popular SQL databases since most of these have support for triggers and procedures.}
@@
 It does not yet support the entire PostgreSQL query syntax.}
-  % Technically speaking this next one is optional.%
-  \item{Adding support for SMOs. These operations allow for more complex schema operations on tables, and should enable software engineers to make more elaborate changes to the database schema.}
+  \item{Explore if Facebook's OSC approach can also be implemented as an alternative migration algorithm, and see where both algorithms differ in performance.}
   \item{Evaluate QuantumDB more extensively in use cases at different organizations and in production environments.}
-\end{itemize}
-
-\section{Early evaluation} %(0.5)
-
-\begin{itemize}
-  \item{Experiment setup}
-  \item{Results}
 \end{itemize}
@@
 \section{Conclusion}
-The conclusion goes here.
+
+Although we are still in the early stages of developing QuantumDB, initial measurements are promising.
+With our small test application we have shown that we can migrate from one database schema to a state where both the old and the new database schemas are active, while still remaining operational.
+Applications can connect with the database according to a specific schema version, and no longer have to deal with database schemas being in an in-between state.
+This effectively decouples the release process from the development process.
 

### 2015-01-23 paper.tex
Added migration state images to illustrate QuantumDB's workings.

@@
 One such example would be adding a new non-nullable column to a PostgreSQL database.
-This operation is blocking in nature --- preventing other queries from running while it is being executed --- and when applied to a table of 100 million rows (in our tests) blocked other queries from running on that table for 2.5 minutes.}
+This operation is blocking in nature --- preventing other queries from running while it is being executed --- and when applied to a table of 50 million rows (in our tests) blocked other queries from running on that table for 2.5 minutes.}
 \end{itemize}
 
-When service windows are unavailable, these two problems make releasing applications often, problematic for release engineers.
+When service windows are unavailable, these two problems make releasing applications frequently, problematic for release engineers.
 To address this problem we aim to find or implement tooling which would allow its users to not only version their database schema, but also evolve their database schema without affecting older and still running applications.
@@
 The previously discussed tools from practice all use approach 1, and prove that this is a viable way for performing simple evolutions of the database schema.
-These tools all create a single ghost table, whereas we aim to create one ghost table for each table that is under change, and ghost tables for all tables which in turn are dependent on these tables under change.
+These tools all create a single ghost table, whereas we aim to create one ghost table for each table that is under change, and ghost tables for all tables which in turn are dependent on these tables under change (see Figure~\ref{fig:migration-stages}).
 This approach ensures that we do not violate any previously defined foreign keys.
-
-\textbf{Show image an example here?}
 
@@
 Although not yet usable in production environments, we have implemented enough features at the moment to take a step back and evaluate QuantumDB in an small experiment.
+
+\begin{figure}%
+    \centering
+    \subfloat[Before migration]{{\includegraphics[width=3.2in]{pre-migration-state.png} }}%
+    \qquad
+    \subfloat[During migration]{{\includegraphics[width=3.2in]{in-migration-state.png} }}%
+    \qquad
+    \subfloat[After migration]{{\includegraphics[width=3.2in]{post-migration-state.png} }}%
+    \caption{Query performance of demo application before and during migration}%
+    \label{fig:migration-stages}%
+\end{figure}
 
@@
 We have installed PostgreSQL 9.3 and a test application --- available on GitHub~\cite{QuantumDB-RelEng-Demo} --- on a server with a 4-core Intel Core i7-3770, a 1TB hard disk, and 16GB memory.
-The test application inserts 100 million records into a single table called ``users'' and starts a simple demo application with two threads which continuously query and manipulate the data in that table using SELECT, INSERT, UPDATE, and DELETE statements.
+The test application inserts 50 million records into a single table called ``users'' and starts a simple demo application with two threads which continuously query and manipulate the data in that table using SELECT, INSERT, UPDATE, and DELETE statements.
 It then uses one of two methods to add a non nullable column to the ``users'' table.

### 2015-01-23 paper.tex
Did a review and rephrasing pass over the paper.

--- 2015-01-21 paper.tex
@@
 %\boldmath
-The abstract goes here.
+
+Continuous Deployment is a hot topic.
+It holds much promise for developing and deploying software applications.
+However both software and release engineers face various challenges when adopting this practice.
+In our view, the biggest technical challenge in the development and deployment of software (which uses SQL databases to store persistent data), is evolving the database schema of SQL databases without disrupting the applications connected to the database.
+In this paper we explore tools, and approaches to address this problem.
 
@@
 SQL databases are an integral part of many (web) applications.
-They provide strong guarantees about the storage and retrieval of data which makes them useful for various software applications.
-However the environment of software engineers and release engineers is drastically changing: shorter development iterations, more frequently deployed to production, or even automatically deployed to production several times a day.
+They provide strong guarantees about the storage and retrieval of persistent data which makes them useful for various software applications.
+However the environment of software and release engineers is drastically changing: shorter development iterations, more frequent deployments to the production environment, or even fully automated deployments to the production environments several times a day.
 For release engineers this brings additional challenges.
-In particular when evolving the schema (tables, columns, foreign keys, etc) of the SQL database either forwards or backwards without loss of data.
+In particular when evolving the database schema (tables, columns, foreign keys, constraints, etc) of the SQL database, or rolling back such changes without loss of data.
 
-This can essentially be distilled into two separate problems:
+We can distill these challenges into two separate problems:
 \begin{itemize}
-  \item{Given the deployment of a new version of a web application, depending on the deployment strategy it might be required that both the old and the new version of the application run in parallel (\textit{Rolling Upgrades}, or \textit{Canary Releases}).
-This means that the schema on the SQL database needs to be compatible with the schema expected by both the old and the new application.}
-  \item{Operations which modify the schema in SQL databases (\textit{DDL statements}), depending on the SQL database implementation may be either non-blocking, or blocking in nature.
-In case of non-blocking DDL statements queries issued by the application for instance may proceed, whereas with blocking DDL statements these queries may be put on hold for the duration of the schema modification.
-This disruption can range anywhere from milliseconds to hours depending on the DDL statement, the size of the affected table, and database implementation.
-One such example would be adding a new non-nullable column to a PostgreSQL database.
-This operation is blocking in nature --- preventing other queries from running while it is being executed --- and when applied to a table of 100 million rows (in our tests) blocked other queries from running on that table for 5 minutes.}
+  \item{Given the deployment of a new version of a (web) application, depending on the deployment strategy it might be required that both the old and the new version of the application run in parallel (\textit{Rolling Upgrades}, or \textit{Canary Releases}~\cite{Humble:2010:CDR:1869904}).
+This means that the schema on the SQL database needs to be compatible with the schema expected by both the old and the new version of the application.}
+  \item{Operations which modify the database schema in a SQL database (\textit{DDL statements}) can be either non-blocking or blocking in nature, depending on the implementation of the operations in that specific SQL database.
+In case of non-blocking \textit{DDL statements}, SQL queries issued by the application may proceed, whereas with blocking \textit{DDL statements} these queries may be put on hold for the duration of the operation.
+This disruption can range from milliseconds to hours depending on the schema operation, the size of the affected table(s), and database implementation.
+One such example would be adding a new non-nullable column to an existing table in a PostgreSQL database.
+This operation is blocking in nature, preventing other queries from running while it is being executed.
+When applied to a table of 50 million records (see Section~\ref{sec:results}), the queries from the application can be blocked for 2.5 minutes.}
 \end{itemize}
 
-When service windows are unavailable, these two problems make releasing applications often, problematic for release engineers.
-To address this problem we aim to find or implement tooling which would allow its users to not only version their database schema, but also evolve their database schema without affecting older and still running applications.
+When service windows are short or even non-existent, these two problems make releasing new versions of applications frequently, problematic for release engineers.
+To address this problem we aim to find or implement a tool which would allow its users to not only alter their database schema in a non-blocking fashion, but also do this without affecting older and still running applications.
 
@@
 
-Before we decided to develop tooling to solve this ourselves, we examined existing approaches, tools, and research which aims to solve these problems.
+We first examined existing approaches, tools, and research which aim to solve these problems.
 A good starting point is the seminal book ``Continuous Delivery''~\cite{Humble:2010:CDR:1869904}.
-This book describes three high-level approaches to deal with migrating data without data loss or requiring downtime:
+This book describes three high-level approaches to deal with migrating data without loss of data or requiring downtime:
 
@@
 Finally replay all the recorded database transactions on the copied database or schema to reach an equivalent state.
-Once this state has been achieved, it must either be maintained by continuing to record and playback transaction, or one must make the switch from the old to the new database or schema.}
-  \item{\textbf{Approach 2}: When using Blue-Green deployments, two instances of the database are available.
-In order to evolve the schema one must apply the schema operations to the inactive database.
+Once this state has been achieved, it must either be maintained by continuing to record and playback transactions, or make the switch from the old to the new database or schema.}
+  \item{\textbf{Approach 2}: When using \textit{Blue-Green deployments}, two instances of the database are available.
+Take the inactive database and apply the schema operations to it.
 In the mean time these two databases need to be kept synchronized.
@@
   \item{\textbf{Approach 3}: The last is an approach where schema operations are performed separately from your application deployments using operations which are non-blocking.
-As a consequence the application must be able to deal with either the old or the new database schema being in use, and formulate queries accordingly.}
+As a consequence the application must be able to deal with either the old or the new database schema being in use, and construct SQL queries which can operate correctly on both schemas.}
 \end{itemize}
@@
 \subsection{Current tooling} 
-Since evolving the schema of a database and coordinating the deployment of software is a complex task, we looked for tooling which could help automate these tasks for developers and release engineers.
+Since evolving the schema of a database and coordinating the deployment of software are both complex tasks, we looked for tooling which could help automate this for software and release engineers.
 
@@
 are tools designed to help DBAs perform administrative tasks on MySQL databases.
-\textit{oak-online-alter-table} and \textit{pt-online-schema-change} respectively are commands which follow a variation of approach 1.
-They create an empty copy of the a specified table and then apply schema operations on the new table.
+\textit{oak-online-alter-table} and \textit{pt-online-schema-change} respectively are commands which create an empty copy of the a specified table (a ghost table) and apply schema operations on ghost table.
 These commands then proceed to install triggers on both tables, which insert, update, or delete the corresponding record in the other table when a record in their own table is inserted, updated, or deleted.
-A background task then iterates over all rows in the original table (in batches) and lazily copies rows from the original table to the new table.
-When all rows have been copied the original table and new table switch names, effectively ensuring that clients now use the new table when querying the old table name.
+A background task iterates over all records in the original table (in batches) and lazily copies records from the original table to the ghost table.
+When all records have been copied the original table and ghost table switch names atomically, ensuring that clients now use the ghost table when querying the original table name.
 
@@
 is a tool which works very similar to Openark Kit and Percona Toolkit.
-But instead of using triggers to synchronize the two tables with each other.
-TableMigrator requires tables to have an additional column which stores the date and time that record was last modified.
-Using this information corresponding rows in both tables are synchronized with each other by passing over all rows in multiple passes.
-Once a pass can be done relatively quickly, TableMigrator acquires an exclusive write lock --- blocking the entire table --- and proceeds to do a final pass.
-Once this has been done, it atomically swaps both table names before releasing the write lock.
+But instead of using triggers to synchronize the two tables with each other,  TableMigrator requires tables to have an additional column which stores the date and time that record was last modified.
+Using this information corresponding records in both tables are synchronized with each other by passing over all records in multiple passes.
+Once a pass can be done relatively quickly, TableMigrator acquires an exclusive write lock on the original table --- blocking the entire table --- and proceeds to do a final pass.
+Once this has been done, it atomically swaps the names of the original and ghost table before releasing the write lock.
+Since this final pass can be done quickly, any application executing queries on that table is only very briefly blocked.
 
@@
 are both tools which closely follow approach 1.
-These tools create a full copy of a table (including its contents) and record modifications to rows in the original table using triggers into a special \textit{deltas} table.
-Once they have copied the original table, they apply the specified schema operations to the new table before proceeding to replay the recorded modifications from the \textit{deltas} table onto the new table.
-When the \textit{deltas} table is almost empty, these tools acquire an exclusive write lock, apply the last recorded modifications and then atomically switch the names of both tables.
+These tools create a copy of a table (including its contents) and record modifications to records in the original table using triggers into a special ``deltas'' table.
+Once they have copied the original table and its contents, they apply the specified schema operations to the ghost table before proceeding to replay the recorded modifications from the ``deltas'' table onto the ghost table.
+When the ``deltas'' table is almost empty, these tools acquire an exclusive write lock on the original table --- like \textit{TableMigrator} does --- and apply the last of the recorded modifications before atomically switch the names of both tables.
 
@@
 
-% Technically we could go without this subsubsection...
-\subsubsection{PRISM}
-PRISM~\cite{Curino:2010:URI:1921071.1921078, Curino:2009:ADS:1656437.1656444} is a tool which is a part of a larger framework.
-Although PRISM does not solve the problem of downtime when performing operations on the database schema, it does aim to shorten it.
-PRISM is designed to reason about the database schema changes, and infer how to rewrite existing queries to the new resulting database schema.
-In addition it can generate the migration and rollback scripts which will transform the current database schema to the new database schema and back.
-The most influential concept of PRISM is the definition of several new operations to evolve a database schema.
-These described Schema Modification Operations (\textit{SMOs}), offer ways to split or merge tables into multiple tables (structurally or content-wise).
-
 \subsubsection{Imago}
 Imago~\cite{Dumitras:2009:WUF:1656980.1657005} is a tool which follows approach 2.
-With Imago there are two production environments --- called "universes" by the authors.
-Imago integrates at both ingress point (the load balancer) to be able to switch users from one universe to the other, and at the egress point where the (web) application sends queries to the SQL database.
-When a new version of the application (requiring a schema change) needs to be deployed, Imago bootstraps the inactive production environment with a copy of the database, applies the required schema operation and then copies over the data from the old universe to the new universe.
-When this process is almost complete, it enforces a short period of read-only access to the SQL database while it is switching users from the old environment to the new, using the load balancer.
+With Imago there are two production environments, called ``universes'' by the authors.
+Imago integrates with the application and its environment at both the ingress point (load balancer) to be able to switch users from one ``universe'' to the other, and at the egress point where the (web) application sends queries to the SQL database.
+When a new version of the application (requiring a change to the database schema) needs to be deployed, Imago bootstraps the inactive production environment with a copy of the active database, applies the required schema operations and then copies over the data from the old universe to the new universe.
+When this process is almost complete, it enforces a short period of read-only access to the SQL database, while it is switching users from the old ``universe'' to the new ``universe'' using the load balancer.
 Imago makes no claims in regards to its support for foreign keys.
@@
   \item{None of the tools used in practice support foreign keys and primarily focus on MySQL databases.}
-  \item{Tools which do not handle the deployment of the application themselves, all force the developers of the application to ensure that their application is able to deal with the database schema being in either the old or the new state.
-All tools atomically rename the modified table, but the software developers of the application are left to deal with modifying queries to accommodate added or removed columns, constraints, or tables at the time of the atomic table rename operation.}
+  \item{These tools are very limited in the sense that they often only support making small changes to the existing database schema.
+This forces software and release engineers to often release intermediate versions of the application, as the schema is altered in several steps.}
 \end{itemize}
@@
 
-We feel that the current situation is an impediment to adoption and growth of \textit{Continuous Deployment}.
-Although \textit{Continuous Deployment} has made great improvements on workflows, team dynamics, and tooling for deploying and monitoring applications, the area of schema evolution in databases has received little attention and no suitable solution currently exists.
-Therefore we aim to address these challenges by developing a new tool which adheres to the following restrictions:
+We feel that the current situation is an impediment to the adoption of \textit{Continuous Deployment}.
+Although \textit{Continuous Deployment} has made great improvements on workflows, team dynamics, and tooling for deploying and monitoring applications, the area of schema evolution in databases has received little attention.
+Therefore we aim to address these problems by developing a new tool which adheres to the following principles:
 
 \begin{itemize}
-  \item{A solution should not put any restrictions on how to deploy the application, or which methods of deployment that can be used: \textit{Blue-Green deployments}, \textit{Rolling upgrades}, \textit{Canary releases}, etc.}
-  \item{The contents of the database must be accessible through one or more database schemas, which are continuously kept synchronized with each other.}
-  \item{Clients connected to the database must be able to ``view'' the contents of the database according to an active database schema of their choice.
-Their ``view'' should only contain tables, columns, foreign keys, etc which are present in their chosen database schema.
-Even when more than one database schema is active at the same time.}
+  \item{A solution should not put restrictions on how to deploy applications, or which methods of deployment can be used: \textit{Blue-Green deployments}, \textit{Rolling upgrades}, \textit{Canary releases}, etc.}
+  \item{Clients connected to the database must be able to access the contents of the database according to an active version of the database schema of their choosing.
+Their ``view'' should only contain tables, columns, foreign keys, etc which are present in their chosen version of the database schema, but contain the same data as is present in all other ``views''.}
   \item{Versioning and rolling back are important for release engineers.
-A solution should support versioning the changes being made to the database schema, and be able to rollback to a previous version.}
+A solution should support versioning the changes being made to the database schema, and be able to rollback to a previous version if the need arises.}
   \item{Foreign keys are an integral part of SQL databases and should be supported.}
@@
 
-We have come up with an approach which combines approaches 1 and 3.
-The previously discussed tools from practice all use approach 1, and prove that this is a viable way for performing simple evolutions of the database schema.
-These tools all create a single ghost table, whereas we aim to create one ghost table for each table that is under change, and ghost tables for all tables which in turn are dependent on these tables under change.
-This approach ensures that we do not violate any previously defined foreign keys.
-
-\textbf{Show image an example here?}
+The previously discussed tools from practice all use approach 1 or a variation of it.
+They prove that this is a viable way for simple evolutions of the database schema.
+However where these tools all create a single ghost table, we aim to create one ghost table for each table that is under change, and ghost tables for all tables which in turn are dependent on these tables under change.
+For instance adding a new non-nullable column ``summary'' to the ``books'' table (see Figure~\ref{fig:migration-stages}), means we should also create a ghost table for the ``publications`` table since there is a foreign key pointing from the ``publications'' table to the ``books'' table.
+The other tables do not require ghost tables, as they do not have any foreign keys referencing either the ``books'' or ``publications'' table.
+This approach ensures that we can support foreign keys.
 
 In addition, the tools from practice all allow release engineers to switch the table names manually after the the original and the ghost table have been synchronized.
-This is good since it allows for two or more versions of the application to co-exist on the same database as long as each version correctly queries the correct (ghost)table.
+This allows for two or more versions of the application to co-exist on the same database as long as each version correctly queries the correct (ghost) table.
 Although the current tools from practice do not solve this specific problem, we think that this is easily solved using a small wrapper around database drivers for most programming languages.
-Alternatively, for software engineers working on the application we could expose which (ghost)table should be queried for a specific version of the database schema, when writing SQL queries.
+
+Finally, with this approach it is possible to perform several schema evolutions at once.
+In the example, we could for instance also add even more columns to either the ``books'' or ``publications'' table.
 
@@
 
-We are implementing a new tool --- QuantumDB~\cite{QuantumDB} --- which aims to address the need for non-blocking schema evolution, and remove obstacles for teams who want to adopt \textit{Continuous Deployment} but struggle because of their SQL database(s).
-Although not yet usable in production environments, we have implemented enough features at the moment to take a step back and evaluate QuantumDB in an small experiment.
+We have started implementing a new tool --- QuantumDB~\cite{QuantumDB} --- which following this approach.
+The aim is to make \textit{Continuous Deployment} more attainable for software and release engineers with respect to applications relying on SQL databases.
+Although not yet ready for production environments, we have implemented enough to perform a quick evaluation in an small experiment.
+
+\begin{figure}%
+    \centering
+    \subfloat[Original database schema, before migration]{{\includegraphics[width=3.0in]{pre-migration-state.png} }}%
+    \qquad
+    \subfloat[Database schema in mixed-state, during migration]{{\includegraphics[width=3.0in]{in-migration-state.png} }}%
+    \qquad
+    \subfloat[New database schema, after migration]{{\includegraphics[width=3.0in]{post-migration-state.png} }}%
+    \caption{Query performance of demo application before and during migration}%
+    \label{fig:migration-stages}%
+\end{figure}
 
@@
   \item{\textbf{core} is the program responsible for performing the actual migration tasks on the database.
-It does this by creating ghost tables of tables under change, using triggers to keep the ghost tables and the original tables synchronized with each other, and a background task which copies the data from the original tables to the ghost tables in small batches to prevent overloading, and degrading the performance of the database.}
+It does this by creating ghost tables of tables under change, using triggers to keep the ghost tables and the original tables synchronized with each other, and a background task which copies the data from the original tables to the ghost tables in small batches to prevent degrading the performance of the database.}
   \item{\textbf{driver} is a Java JDBC driver which can wrap another JDBC driver.
-When used, the application developers can specify a schema version on which their application operates.
-The driver then rewrites SQL queries sent to it by the application, and passes them on to the wrapped database JDBC driver which in turn sends it to the database.
-When rewriting queries, it replaces table names with the names of the associated (ghost)table names if applicable.}
+When used, the software engineers can specify a database schema version on which that version of their application operates.
+The driver then rewrites SQL queries sent to it by the application, and passes them on to the wrapped database JDBC driver which in turn sends them to the database.
+When rewriting queries, it replaces table names with the names of the associated (ghost) table names where applicable.}
 \end{itemize}
 
-With QuantumDB we can currently take one of the active database schemas, apply an arbitrary number of schema operations to it, and expose the resulting database schema in addition to the original database schemas.
-Once none of the applications are still connected to a particular version of the database schema, it can safely be discarded in a background task.
-Effectively this can be described as A/B testing the database schema.
+With QuantumDB we can currently take one of the active database schemas, apply an arbitrary number of schema operations to it, and expose the resulting database schema in addition to the original database schema.
+Once the last application connected to a particular version of the database schema disconnects, the tables associated with that database schema version can be safely discarded.
 
-\subsection{Road ahead}
+\section{Initial measurements} %(0.5)
 
-As stated before, the implementation of QuantumDB is not yet ready for production. To be able to use QuantumDB effectively in production environments, we still need to implement some features.
+We feel it is important to validate our work at an early stage.
+Although the development of QuantumDB is still ongoing, we setup a small experiment to verify that the concept and our implementation of it works.
+
+\subsection{Setup}
+We have installed PostgreSQL 9.3 and a small test application --- made available on GitHub~\cite{QuantumDB-RelEng-Demo} --- on a server with a 4-core Intel Core i7-3770, a 1TB hard disk, and 16GB memory.
+The test application inserts 50 million records into a single table called ``users'' and starts a simple demo application with two threads which continuously query and manipulate the data in that table using SELECT, INSERT, UPDATE, and DELETE statements.
+It then uses one of two methods to add a non nullable column to the ``users'' table.
+During this process we keep track of the duration of the queries issued by the demo application and how long each method takes to complete.
+
+The ``naive'' approach simply executes an ``ALTER TABLE users ADD COLUMN'' statement to add the new column to the database.
+Once this statement has completed, we can start a new demo application which can operate on the new structure of the ``users'' table.
+Since the new structure of the ``users'' table is compatible with both the old and the new version of the demo application they can run in parallel to each other, allowing for any kind of deployment method to be used.
+
+The second approach uses QuantumDB to add the new column.
+It first creates a ghost table based on the ``users'' table with the new column.
+It then iteratively copies records from the ``users'' table to the ghost table in batches of 2,000 records at intervals of 50 milliseconds.
+With this we attempt to avoid negatively affecting the performance of the queries issued by the demo application while we are copying the records.
+Unlike with the previous approach, with this approach the demo applications use the QuantumDB JDBC driver which expose each demo application to a different version of the database schema (containing either the ``users'' table or the ghost table).
+The driver ensures that each query is rewritten in such a way that the correct table is queried for that specific version of the schema.
+
+\subsection{Results}
+\label{sec:results}
+
+The total migration time for the ``naive'' approach was 2 minutes and 26 seconds, during which the demo application was unable to read from or write to the database.
+The total migration time for the QuantumDB approach was 50 minutes and 41 seconds.
+Although significantly slower than the ``naive'' approach, the demo applications remained operational throughout this period (see Figure~\ref{fig:migration}).
+
+\begin{figure}%
+    \centering
+    \subfloat[Before migration]{{\includegraphics[width=1.55in]{normal-operations.png} }}%
+    \qquad
+    \subfloat[During migration]{{\includegraphics[width=1.55in]{migration.png} }}%
+    \caption{Query performance of demo application before and during migration}%
+    \label{fig:migration}%
+\end{figure}
+
+It is worth noting that we have not invested any time in improving the performance of QuantumDB or reducing the performance impact it has on the database and its clients.
+During this experiment we noticed that the hard disk was periodically unable to write data to disk quickly enough.
+As a result queries were taking significantly longer than normal to complete.
+We are confident that we can greatly reduce this by tweaking the PostgreSQL config, or switching to a faster hard disk.
+
+\section{Road ahead}
+As stated before, work on QuantumDB is not finished. To be able to use QuantumDB effectively in production environments, we still need to work on the following issues:
 
@@
   \item{QuantumDB can already deal with foreign keys referring from tables under change to other tables not under change.
-To fully support foreign keys QuantumDB would also need to be able to deal with foreign keys referring to tables under change.}
-  \item{QuantumDB currently only supports PostgreSQL, but should prove easily extensible to other popular SQL databases since most of these have support for triggers and procedures.}
+To fully support foreign keys QuantumDB would also need to be able to handle foreign keys referring to tables under change.
+This is currently being worked on.}
+  \item{Use performance metrics to throttle the copy phase.
+When the database is (too) busy, (high disk I/O, high CPU usage, or memory starvation), we should slow down the iterative copy process.
+Vice-versa we can speed it up when the database is not under load.}
+  \item{QuantumDB currently only supports PostgreSQL, but since most SQL databases have support for triggers and procedures of some kind, this should prove easily extendable to other popular SQL databases.}
   \item{The driver currently uses a very rudimentary query rewriting algorithm to replace table names under change with the names of ghost tables.
 It does not yet support the entire PostgreSQL query syntax.}
-  % Technically speaking this next one is optional.%
-  \item{Adding support for SMOs. These operations allow for more complex schema operations on tables, and should enable software engineers to make more elaborate changes to the database schema.}
-  \item{Evaluate QuantumDB more extensively in use cases at different organizations and in production environments.}
-\end{itemize}
-
-\section{Early evaluation} %(0.5)
-
-\begin{itemize}
-  \item{Experiment setup}
-  \item{Results}
 \end{itemize}
@@
 \section{Conclusion}
-The conclusion goes here.
+
+Although development of QuantumDB is still ongoing, initial measurements are promising.
+With our small test application we have shown that we can migrate from one database schema to a state where both the old and the new database schemas are active, while still enabling database clients to remain operational.
+Applications can connect with the database according to a specific schema version, and no longer have to deal with database schemas being in an in-between state.
+This should effectively allow further decoupling of the release process from the development process.
+We believe that this approach can put \textit{Continuous Deployment} and \textit{Canary Releases} in reach of more software and release engineers.
 

### 2015-01-23 paper.tex
Removed some unnecessary commas.

@@
 \section{Road ahead}
-As stated before, work on QuantumDB is not finished. To be able to use QuantumDB effectively in production environments, we still need to work on the following issues:
+As stated before, work on QuantumDB is not finished. To be able to use QuantumDB effectively in production environments we still need to work on the following issues:
 
@@
   \item{Use performance metrics to throttle the copy phase.
-When the database is (too) busy, (high disk I/O, high CPU usage, or memory starvation), we should slow down the iterative copy process.
+When the database is (too) busy (high disk I/O, high CPU usage, or memory starvation), we should slow down the iterative copy process.
 Vice-versa we can speed it up when the database is not under load.}

### 2015-01-29 paper.tex
Small textual changes to sections I-III.

@@
 They provide strong guarantees about the storage and retrieval of persistent data which makes them useful for various software applications.
-However the environment of software and release engineers is drastically changing: shorter development iterations, more frequent deployments to the production environment, or even fully automated deployments to the production environments several times a day.
-For release engineers this brings additional challenges.
-In particular when evolving the database schema (tables, columns, foreign keys, constraints, etc) of the SQL database, or rolling back such changes without loss of data.
 
-We can distill these challenges into two separate problems:
-\begin{itemize}
-  \item{Given the deployment of a new version of a (web) application, depending on the deployment strategy it might be required that both the old and the new version of the application run in parallel (\textit{Rolling Upgrades}, or \textit{Canary Releases}~\cite{Humble:2010:CDR:1869904}).
-This means that the schema on the SQL database needs to be compatible with the schema expected by both the old and the new version of the application.}
-  \item{Operations which modify the database schema in a SQL database (\textit{DDL statements}) can be either non-blocking or blocking in nature, depending on the implementation of the operations in that specific SQL database.
+For release engineers, the persistence of the data and especially the structure of this data as embedded in the application's SQL schema is at odds with today's need for continuous deployment.
+Challenges include combining zero downtime constraints with evolving database schemas (affecting tables, columns, foreign keys, constraints, etc.), and the need to be able to rollback changes without loss of data.
+
+We can distill the challenge of continuous delivery in the presence of schema evolution into two separate problems:
+
+First, given the deployment of a new version of a (web) application, depending on the deployment strategy it might be required that both the old and the new version of the application run in parallel (\textit{Rolling Upgrades}, or \textit{Canary Releases}~\cite{Humble:2010:CDR:1869904}).
+This means that the schema on the SQL database needs to be compatible with the schema expected by both the old and the new version of the application.
+
+Second, operations which modify the database schema in a SQL database (\textit{DDL statements}) can be either non-blocking or blocking in nature, depending on the implementation of the operations in that specific SQL database.
 In case of non-blocking \textit{DDL statements}, SQL queries issued by the application may proceed, whereas with blocking \textit{DDL statements} these queries may be put on hold for the duration of the operation.
-This disruption can range from milliseconds to hours depending on the schema operation, the size of the affected table(s), and database implementation.
+This disruption can range from milliseconds to hours depending on the schema operation, the size of the affected table(s), and the database implementation.
 One such example would be adding a new non-nullable column to an existing table in a PostgreSQL database.
 This operation is blocking in nature, preventing other queries from running while it is being executed.
-When applied to a table of 50 million records (see Section~\ref{sec:results}), the queries from the application can be blocked for 2.5 minutes.}
-\end{itemize}
+When applied to a table of 50 million records (see Section~\ref{sec:results}), the queries from the application can be blocked for 2.5 minutes.
 
-When service windows are short or even non-existent, these two problems make releasing new versions of applications frequently, problematic for release engineers.
-To address this problem we aim to find or implement a tool which would allow its users to not only alter their database schema in a non-blocking fashion, but also do this without affecting older and still running applications.
+When service windows are short or even non-existent, these two problems make releasing new versions of applications frequently problematic for release engineers.
+To address this problem we aim to find or implement a tool which  allows its users to not only alter their database schema in a non-blocking fashion, but also do this without affecting older and still running applications.
 
@@
 
-We first examined existing approaches, tools, and research which aim to solve these problems.
-A good starting point is the seminal book ``Continuous Delivery''~\cite{Humble:2010:CDR:1869904}.
+\subsection{Continuous Delivery}\label{sec:cd}
+
+We first examine existing approaches, tools, and research aiming to solve these problems.
+Our starting point is the seminal book ``Continuous Delivery''~\cite{Humble:2010:CDR:1869904}.
 This book describes three high-level approaches to deal with migrating data without loss of data or requiring downtime:
@@
 This can be done by using approach 1, or somehow ensure that the table under change remains in a read-only state until the active and inactive databases have switched roles.}
-  \item{\textbf{Approach 3}: The last is an approach where schema operations are performed separately from your application deployments using operations which are non-blocking.
+  \item{\textbf{Approach 3}: The last is an approach where schema operations are performed separately from the application deployments using operations which are non-blocking.
 As a consequence the application must be able to deal with either the old or the new database schema being in use, and construct SQL queries which can operate correctly on both schemas.}
@@
 \subsubsection{Percona Toolkit~\cite{PerconaToolkit} \& Openark Kit~\cite{OpenarkKit}}
-are tools designed to help DBAs perform administrative tasks on MySQL databases.
-\textit{oak-online-alter-table} and \textit{pt-online-schema-change} respectively are commands which create an empty copy of the a specified table (a ghost table) and apply schema operations on ghost table.
+These are tools designed to help DBAs perform administrative tasks on MySQL databases.
+The commands \textit{oak-online-alter-table} and \textit{pt-online-schema-change} create an empty copy of the a specified table (a ghost table) and apply schema operations to the ghost table.
 These commands then proceed to install triggers on both tables, which insert, update, or delete the corresponding record in the other table when a record in their own table is inserted, updated, or deleted.
@@
 \subsubsection{TableMigrator~\cite{TableMigrator}}
-is a tool which works very similar to Openark Kit and Percona Toolkit.
-But instead of using triggers to synchronize the two tables with each other,  TableMigrator requires tables to have an additional column which stores the date and time that record was last modified.
+This is a tool which works like Openark Kit and Percona Toolkit.
+Instead of using triggers to synchronize the two tables with each other,  TableMigrator requires tables to have an additional column which stores the date and time that record was last modified.
 Using this information corresponding records in both tables are synchronized with each other by passing over all records in multiple passes.
@@
 \subsubsection{SoundCloud's Large Hadron Migrator~\cite{SoundcloudLHM} \& Facebook's Online Schema Change~\cite{FacebookOSC}}
-are both tools which closely follow approach 1.
-These tools create a copy of a table (including its contents) and record modifications to records in the original table using triggers into a special ``deltas'' table.
+These are both tools which closely follow Approach~1 descriged earlier (Section~\ref{sec:cd}).
+They create a copy of a table (including its contents) and record modifications to records in the original table using triggers into a special ``deltas'' table.
 Once they have copied the original table and its contents, they apply the specified schema operations to the ghost table before proceeding to replay the recorded modifications from the ``deltas'' table onto the ghost table.
@@
 \subsubsection{Imago}
-Imago~\cite{Dumitras:2009:WUF:1656980.1657005} is a tool which follows approach 2.
-With Imago there are two production environments, called ``universes'' by the authors.
+Imago~\cite{Dumitras:2009:WUF:1656980.1657005} is a tool which follows Approach~2.
+With Imago there are two production environments, called ``universes''.
 Imago integrates with the application and its environment at both the ingress point (load balancer) to be able to switch users from one ``universe'' to the other, and at the egress point where the (web) application sends queries to the SQL database.
@@
 \begin{itemize}
-  \item{None of the tools used in practice support foreign keys and primarily focus on MySQL databases.}
-  \item{These tools are very limited in the sense that they often only support making small changes to the existing database schema.
-This forces software and release engineers to often release intermediate versions of the application, as the schema is altered in several steps.}
+  \item None of the tools used in practice support foreign keys
+  \item The tools primarily focus on MySQL databases.
+  \item These tools are limited in the sense that they typically only support making small changes to the existing database schema.
+This forces software and release engineers to often release intermediate versions of the application, as the schema is altered in several steps.
 \end{itemize}
 
-\section{Our approach} %(1.0)
+\section{Proposed Approach} %(1.0)
 
@@
 
-The previously discussed tools from practice all use approach 1 or a variation of it.
+The previously discussed tools from practice all use Approach~1 or a variation of it.
 They prove that this is a viable way for simple evolutions of the database schema.

### 2015-01-29 paper.tex
Small fixes seciton IV-VII

@@
     Delft University of Technology,\\
-    Delft, Netherlands\\ \\
+    Delft, The Netherlands\\ \\
     Email: M.deJong-2@student.tudelft.nl\IEEEauthorrefmark{1}\\
@@
 
-\section{State of the art} %(0.7)
+\section{State of the Art} %(0.7)
 
@@
 
-\subsection{Current tooling} 
+\subsection{Current Tooling} 
 Since evolving the schema of a database and coordinating the deployment of software are both complex tasks, we looked for tooling which could help automate this for software and release engineers.
@@
 
-\subsection{Solution direction}
+\subsection{Solution Direction}
 
@@
 
-\section{Implementation status} %(0.5)
+\section{Implementation Status} %(0.5)
 
-\subsection{Current status}
+\subsection{Current Status}
 
-We have started implementing a new tool --- QuantumDB~\cite{QuantumDB} --- which following this approach.
-The aim is to make \textit{Continuous Deployment} more attainable for software and release engineers with respect to applications relying on SQL databases.
-Although not yet ready for production environments, we have implemented enough to perform a quick evaluation in an small experiment.
+We are in the process of implementing and experimenting with the approach just described via a new tool that we have named QuantumDB~\cite{QuantumDB}.
+The aim is to make \textit{Continuous Deployment} more attainable for software and release engineers for applications relying on SQL databases.
+Although not yet ready for production environments, we have implemented enough to perform a quick evaluation in a small experiment.
 
@@
 \begin{itemize}
-  \item{\textbf{core} is the program responsible for performing the actual migration tasks on the database.
+  \item{\textbf{qdb:core} is the program responsible for performing the actual migration tasks on the database.
 It does this by creating ghost tables of tables under change, using triggers to keep the ghost tables and the original tables synchronized with each other, and a background task which copies the data from the original tables to the ghost tables in small batches to prevent degrading the performance of the database.}
-  \item{\textbf{driver} is a Java JDBC driver which can wrap another JDBC driver.
+  \item{\textbf{qdb:driver} is a Java JDBC driver which can wrap another JDBC driver.
 When used, the software engineers can specify a database schema version on which that version of their application operates.
@@
 
-\section{Initial measurements} %(0.5)
+\section{Initial Measurements} %(0.5)
 
-We feel it is important to validate our work at an early stage.
-Although the development of QuantumDB is still ongoing, we setup a small experiment to verify that the concept and our implementation of it works.
+In order to validate our work at an early stage we have setup a small experiment to verify that the proposed approach and our implementation of it behaves as expected.
 
@@
 The test application inserts 50 million records into a single table called ``users'' and starts a simple demo application with two threads which continuously query and manipulate the data in that table using SELECT, INSERT, UPDATE, and DELETE statements.
-It then uses one of two methods to add a non nullable column to the ``users'' table.
+It then uses one of two methods to add a non-nullable column to the ``users'' table.
 During this process we keep track of the duration of the queries issued by the demo application and how long each method takes to complete.
@@
 With this we attempt to avoid negatively affecting the performance of the queries issued by the demo application while we are copying the records.
-Unlike with the previous approach, with this approach the demo applications use the QuantumDB JDBC driver which expose each demo application to a different version of the database schema (containing either the ``users'' table or the ghost table).
+Unlike with the previous approach, with this approach the demo applications use the QuantumDB JDBC driver which exposes each demo application to a different version of the database schema (containing either the ``users'' table or the ghost table).
 The driver ensures that each query is rewritten in such a way that the correct table is queried for that specific version of the schema.

### 2015-01-29 paper.tex
Rewrote abstract.

@@
 
-Continuous Deployment is a hot topic.
-It holds much promise for developing and deploying software applications.
-However both software and release engineers face various challenges when adopting this practice.
-In our view, the biggest technical challenge in the development and deployment of software (which uses SQL databases to store persistent data), is evolving the database schema of SQL databases without disrupting the applications connected to the database.
-In this paper we explore tools, and approaches to address this problem.
+Continuous Deployment is an important enabler rapid delivery of business value and early end user feedback.
+While frequent code deployment is well understood, the impact of frequent change on persistent data is less understood and supported.
+SQL schema evolutions in particular can make it expensive to deploy a new version, and may even lead to downtime if schema changes can only be applied by blocking operations.
+
+In this paper we study the problem of continuous deployment in the presence of database schema evolution in more detail.
+We identify a number of shortcomings to existing solutions and tools, mostly related to avoidable down time and support for foreign keys.
+We propose a novel approach to address these problems, and provide an open source implementation.
+Initial evaluation suggests the approach is effective and sufficiently efficient.
 

### 2015-01-29 paper.tex
Fixed minor typos and rephrased some sentences.

@@
 
-Continuous Deployment is an important enabler rapid delivery of business value and early end user feedback.
+Continuous Deployment is an important enabler of rapid delivery of business value and early end user feedback.
 While frequent code deployment is well understood, the impact of frequent change on persistent data is less understood and supported.
@@
 In this paper we study the problem of continuous deployment in the presence of database schema evolution in more detail.
-We identify a number of shortcomings to existing solutions and tools, mostly related to avoidable down time and support for foreign keys.
+We identify a number of shortcomings to existing solutions and tools, mostly related to avoidable downtime and support for foreign keys.
 We propose a novel approach to address these problems, and provide an open source implementation.
@@
 When service windows are short or even non-existent, these two problems make releasing new versions of applications frequently problematic for release engineers.
-To address this problem we aim to find or implement a tool which  allows its users to not only alter their database schema in a non-blocking fashion, but also do this without affecting older and still running applications.
+To address this problem we aim to find or implement a tool which allows its users to not only alter their database schema in a non-blocking fashion, but also do this without affecting older applications still operating on the older version of the schema.
 
@@
 \subsubsection{SoundCloud's Large Hadron Migrator~\cite{SoundcloudLHM} \& Facebook's Online Schema Change~\cite{FacebookOSC}}
-These are both tools which closely follow Approach~1 descriged earlier (Section~\ref{sec:cd}).
+These are both tools which closely follow Approach~1 described earlier (Section~\ref{sec:cd}).
 They create a copy of a table (including its contents) and record modifications to records in the original table using triggers into a special ``deltas'' table.
@@
 \begin{itemize}
-  \item None of the tools used in practice support foreign keys
+  \item None of the tools used in practice support foreign keys.
   \item The tools primarily focus on MySQL databases.
@@
 As a result queries were taking significantly longer than normal to complete.
-We are confident that we can greatly reduce this by tweaking the PostgreSQL config, or switching to a faster hard disk.
+We are confident that we can greatly reduce this by tweaking the PostgreSQL config, switching to a faster hard disk, or lowering the rate of migration during the copy phase.
 

### 2015-02-22 paper.tex
Preparing camera-ready version by listing feedback.
Processed some of the feedback provided by reviewers.

@@
   \IEEEauthorblockA{
-    Delft University of Technology,\\
-    Delft, The Netherlands\\ \\
-    Email: M.deJong-2@student.tudelft.nl\IEEEauthorrefmark{1}\\
+    Delft University of Technology,
+    Delft, The Netherlands\\
+    Email: M.deJong-2@student.tudelft.nl\IEEEauthorrefmark{1}
     Email: Arie.vanDeursen@tudelft.nl\IEEEauthorrefmark{2}\\ \\
@@
   \item None of the tools used in practice support foreign keys.
-  \item The tools primarily focus on MySQL databases.
+  \item The tools discussed primarily focus on MySQL databases. \textit{OpenArk Kit} and \textit{Percona Toolkit} are suits of tools which focus exclusively on MySQL database administration.
   \item These tools are limited in the sense that they typically only support making small changes to the existing database schema.
@@
 The previously discussed tools from practice all use Approach~1 or a variation of it.
-They prove that this is a viable way for simple evolutions of the database schema.
+They prove that Approach~1 is a viable way for small evolutions of the database schema.
 However where these tools all create a single ghost table, we aim to create one ghost table for each table that is under change, and ghost tables for all tables which in turn are dependent on these tables under change.
@@
 The other tables do not require ghost tables, as they do not have any foreign keys referencing either the ``books'' or ``publications'' table.
-This approach ensures that we can support foreign keys.
 
@@
 
+Another difference worth mentioning is that in our approach the tables never actually switch names.
+Ultimately when making the switch we simply delete the original table (and its dependencies), and continue to read from and write to the ghost table.
+
 Finally, with this approach it is possible to perform several schema evolutions at once.
@@
 \section{Implementation Status} %(0.5)
-
-\subsection{Current Status}
 
@@
     \centering
-    \subfloat[Original database schema, before migration]{{\includegraphics[width=3.0in]{pre-migration-state.png} }}%
+    \subfloat[Original database schema]{{\includegraphics[width=2.6in]{pre-migration-state.png} }}%
     \qquad
-    \subfloat[Database schema in mixed-state, during migration]{{\includegraphics[width=3.0in]{in-migration-state.png} }}%
+    \subfloat[Database schema in mixed-state]{{\includegraphics[width=2.6in]{in-migration-state.png} }}%
     \qquad
-    \subfloat[New database schema, after migration]{{\includegraphics[width=3.0in]{post-migration-state.png} }}%
-    \caption{Query performance of demo application before and during migration}%
+    \subfloat[New database schema]{{\includegraphics[width=2.6in]{post-migration-state.png} }}%
+    \caption{Different stages of the database schema when modifying the ``books'' table.}%
     \label{fig:migration-stages}%
@@
   \item{\textbf{qdb:core} is the program responsible for performing the actual migration tasks on the database.
-It does this by creating ghost tables of tables under change, using triggers to keep the ghost tables and the original tables synchronized with each other, and a background task which copies the data from the original tables to the ghost tables in small batches to prevent degrading the performance of the database.}
+It does this by creating ghost tables of tables under change, using triggers to keep the ghost tables and the original tables synchronized with each other, and a background task which copies the data from the original tables to the ghost tables in small batches to prevent degrading the performance of the database.
+We know that for PostgreSQL triggers are executed within the same transaction as what triggered the trigger.
+In our case it means that any modification to the original tables is also performed on the ghost tables in the same transaction.}
   \item{\textbf{qdb:driver} is a Java JDBC driver which can wrap another JDBC driver.
@@
 With this we attempt to avoid negatively affecting the performance of the queries issued by the demo application while we are copying the records.
-Unlike with the previous approach, with this approach the demo applications use the QuantumDB JDBC driver which exposes each demo application to a different version of the database schema (containing either the ``users'' table or the ghost table).
+With this approach the demo applications use the QuantumDB JDBC driver which exposes each demo application to a different version of the database schema.
 The driver ensures that each query is rewritten in such a way that the correct table is queried for that specific version of the schema.
@@
     \subfloat[During migration]{{\includegraphics[width=1.55in]{migration.png} }}%
-    \caption{Query performance of demo application before and during migration}%
+    \caption{Query performance before and during migration}%
     \label{fig:migration}%
@@
 
-It is worth noting that we have not invested any time in improving the performance of QuantumDB or reducing the performance impact it has on the database and its clients.
+We have not invested any time in improving the performance of QuantumDB or reducing the performance impact it has on the database and its clients.
 During this experiment we noticed that the hard disk was periodically unable to write data to disk quickly enough.
@@
 \begin{itemize}
-  \item{QuantumDB can already deal with foreign keys referring from tables under change to other tables not under change.
-To fully support foreign keys QuantumDB would also need to be able to handle foreign keys referring to tables under change.
-This is currently being worked on.}
+  \item{QuantumDB can already deal with foreign keys referring from tables under change to tables not under change.
+To fully support foreign keys QuantumDB would also need to be able to handle foreign keys referring to tables under change.}
   \item{Use performance metrics to throttle the copy phase.
@@
   \item{QuantumDB currently only supports PostgreSQL, but since most SQL databases have support for triggers and procedures of some kind, this should prove easily extendable to other popular SQL databases.}
-  \item{The driver currently uses a very rudimentary query rewriting algorithm to replace table names under change with the names of ghost tables.
-It does not yet support the entire PostgreSQL query syntax.}
+  \item{The driver currently uses a very naive method of query rewriting to replace table names under change with the names of ghost tables. The PostgreSQL query syntax is only partially supported.}
 \end{itemize}
@@
 
-Although development of QuantumDB is still ongoing, initial measurements are promising.
+Development of QuantumDB is still ongoing, but initial measurements are promising.
 With our small test application we have shown that we can migrate from one database schema to a state where both the old and the new database schemas are active, while still enabling database clients to remain operational.

### 2015-02-23 paper.tex
Processed last of the feedback.

@@
 
-When service windows are short or even non-existent, these two problems make releasing new versions of applications frequently problematic for release engineers.
+When service windows are short or non-existent, these two problems make releasing new versions of applications frequently problematic for release engineers.
 To address this problem we aim to find or implement a tool which allows its users to not only alter their database schema in a non-blocking fashion, but also do this without affecting older applications still operating on the older version of the schema.
@@
 In the mean time these two databases need to be kept synchronized.
-This can be done by using approach 1, or somehow ensure that the table under change remains in a read-only state until the active and inactive databases have switched roles.}
+This can be done by using Approach 1, or somehow ensure that the table under change remains in a read-only state until the active and inactive databases have switched roles.}
   \item{\textbf{Approach 3}: The last is an approach where schema operations are performed separately from the application deployments using operations which are non-blocking.
@@
 \subsubsection{TableMigrator~\cite{TableMigrator}}
-This is a tool which works like Openark Kit and Percona Toolkit.
+This tool works like Openark Kit and Percona Toolkit.
 Instead of using triggers to synchronize the two tables with each other,  TableMigrator requires tables to have an additional column which stores the date and time that record was last modified.
@@
   \item None of the tools used in practice support foreign keys.
+These tools will not work when a foreign key is present on a table under change.
+This forces software engineers to choose between not using any of these tools, or deliberately avoid defining foreign key constraints in their SQL databases in order to be able to use these tools.
   \item The tools discussed primarily focus on MySQL databases. \textit{OpenArk Kit} and \textit{Percona Toolkit} are suits of tools which focus exclusively on MySQL database administration.
@@
   \item{Clients connected to the database must be able to access the contents of the database according to an active version of the database schema of their choosing.
-Their ``view'' should only contain tables, columns, foreign keys, etc which are present in their chosen version of the database schema, but contain the same data as is present in all other ``views''.}
+Their ``view'' should only contain tables, columns, foreign keys, etc which are present in their chosen version of the database schema.}
   \item{Versioning and rolling back are important for release engineers.
@@
     \centering
-    \subfloat[Original database schema]{{\includegraphics[width=2.6in]{pre-migration-state.png} }}%
+    \subfloat[Original database schema]{{\includegraphics[width=2.75in]{pre-migration-state.png} }}%
     \qquad
-    \subfloat[Database schema in mixed-state]{{\includegraphics[width=2.6in]{in-migration-state.png} }}%
+    \subfloat[Database schema in mixed-state]{{\includegraphics[width=2.75in]{in-migration-state.png} }}%
     \qquad
-    \subfloat[New database schema]{{\includegraphics[width=2.6in]{post-migration-state.png} }}%
+    \subfloat[New database schema]{{\includegraphics[width=2.75in]{post-migration-state.png} }}%
     \caption{Different stages of the database schema when modifying the ``books'' table.}%
